---
title: "Week 3 seminar: more regression, some causality"
author: Joshua Loftus
output: html_document
---

```{r setup, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
theme_set(theme_minimal(base_size = 14))
set.seed(1)
```
## What is a Causal Inference Problem?
A causal inference problem is one in which the goal is to determine the causal effect of an intervention or treatment on an outcome. Unlike simple statistical associations, causal inference aims to answer questions like "Does X cause Y?", rather than just describing correlations. For instance, a causal inference question might be "Does increasing advertising budget lead to higher sales?". The answer YES may seem convincing, but higher sales may have credit to various factors. For example, ice cream sells more in summer, regardless of advertising.

## Regression models with simulated data

### Direct causation `X -> Y`
In this example, we can imagine X: Advertising spend and Y: Sales.

#### Causal model

This describes how the simulated world works.

```{r}
# Direct causation: X -> Y
n <- 200
X <- rnorm(n, mean = 1)
Y_CEF <- function(X) 2 * X
Y <- Y_CEF(X) + rnorm(n)
```
$$Y=2X+U$$
Random noise is independent of X, meaning there is no confounding variable affecting both X and Y. This makes correlation between X and Y reflects the causation between them.

```{r}
# or GGally::ggcoef() for a plot
lm(Y ~ X) |> broom::tidy() |> knitr::kable()
```


#### Counterfactual values for `X`

Below are several examples of different kinds of interventions. We can use these to ask various questions like, "What if the value of `X` had been different in this particular way? Would that have changed other variables as well?"

```{r}
#X_new <- rep(0, n) # "atomic" intervention, set all X to a constant, e.g. 0
X_new <- X + 1 # constant additive intervention
#X_new <- rnorm(n, mean = 2) # shift in mean
```

#### Propagating the intervention

After intervening on a variable we have to re-generate all the other variables that depend on it using its new values. Usually we call the variable with intervention Treatment group and the other Control group, denoted by $Y(1)$ and $Y(0)$.

```{r}
Y_new <- Y_CEF(X_new) + rnorm(n)
```

#### Measuring the causal effect
$$ATE = \mathbb{E}[Y(1)]-\mathbb{E}[Y(0)]$$
In this case, 
$$Y(1)=X_{new}=X+1~~;~~ Y(0)=X $$ 
Randomized Controlled Trials (RCTs): In an ideal setting, ATE is estimated using randomized experiments where treatment assignment is random. This ensures that the treatment and control groups are comparable. In this scenario,
$$\hat{ATE}=\frac{1}{N_T}\sum_{i\in Treatment}Y_i(1)-\frac{1}{N_C}\sum_{i\in Control}Y_i(0)$$
Observational Studies: When randomized experiments aren’t possible, observational data is used.

```{r}
# Average causal effect (ATE)
mean(Y_new - Y)
```

There is another possible treatment effect we may be interested in. Sometimes the magnitude/direction of the treatment effect depends in an important way on another variable, like one of the predictors or covariates. In cases like this we may want to know how the treatment effect varies as we look at different values of the covariate, as in the plot below.

```{r message=FALSE}
# Conditional average treatment effect (CATE)
qplot(X, Y_new - Y, geom=c("smooth", "point")) +
  geom_hline(yintercept = mean(Y_new - Y))
```
In the above plot, variable X is conditioned on different values and we can observe various ATE according to various values of X.

**Exercise**: Can you change `Y_CEF` so that the regression estimate is highly misleading?

**Exercise**: Can you make the `ATE` and the `CATE` provide very different conclusions about the causal effect of `X`?

The following is a concrete example answering the above questions:
```{r}
# Generate data
n <- 500
X <- rnorm(n, mean = 0) 
epsilon <- rnorm(n, sd = 0.5)  
Y_CEF <- function(X) ifelse(X > 0, 2 * X^2, -3 * X) 
Y <- Y_CEF(X) + epsilon
X_new <- X + 1 
Y_new <- Y_CEF(X_new) + epsilon  
ggplot(data.frame(X, Y_diff = Y_new - Y), aes(x = X, y = Y_diff)) +
  geom_point() +
  geom_smooth(method = "loess", color = "blue") +  
  geom_hline(yintercept = mean(Y_new - Y), linetype = "dashed", color = "red") +  
  labs(title = "CATE and ATE: How the Treatment Effect Varies by X",
       x = "X (Covariate)", y = "Treatment Effect (Y_new - Y)") +
  theme_minimal()

ATE <- mean(Y_new - Y)
CATE_positive <- mean(Y_new[X > 0] - Y[X > 0])  # CATE for X > 0
CATE_negative <- mean(Y_new[X <= 0] - Y[X <= 0])  # CATE for X <= 0
```
When $X>0$, $Y=2*X^2$, $Y_{new}=2*(X+1)^2$, $Y_{new}-Y=4X+2$, which is linear in $X$, and it does not aligh with ATE; When $X<0$, $Y=-3*X$, $Y_{new}=-3*(X+1)$, $Y_{new}-Y=-3$, which does not vary with $X$, but it does not equal to ATE. This results in very different ATE and CATE. And on the domain $X>0$, the function is quadratic instead of linear, this leads to non-constant (if we exclude random noise) CATE, which means linear regression estimate is misleading, because the intervention is constant additive, if the relationship is linear, the CATE should be constant.


### Unmeasured confounder `X <- U -> Y
X: Ice cream sales
Y: Number of drownings
U: Temperature`

#### Causal model

In this world there is no causal relationship between `X` and `Y`, because ice cream sales does not affect the number of drownings.

```{r}
n <- 1000
U <- rnorm(n)
X <- -2 * U + U^2 + rnorm(n)
Y_CEF <- function(U) 4 - U
Y <- Y_CEF(U) + rnorm(n)
```

But if we fit regression models we might believe there is a relationship:

```{r}
lm(Y ~ X) |> broom::tidy() |> knitr::kable()
```

And someone looking at a plot might conclude there is a clear, strong relationship:

```{r}
qplot(X, Y)
```

#### Intervening on `X`

```{r}
# Try changing this
X_new <- X + 1
```

#### Propagating the intervention

In this model `X <- U -> Y` there are no other variables on directed pathways out of `X` so nothing else is changed.

```{r}
U_new <- U
Y_new <- Y_CEF(U_new) + rnorm(n) 
```

#### Measuring the causal effect

All causal effects on `Y` are 0 (just random noise around 0)

```{r}
# ATE
mean(Y_new - Y)
```

```{r message=FALSE}
# CATE
qplot(X, Y_new - Y, geom=c("point")) 
```

### Confounded relationship `X -> Y <- Z` and `Z -> X`

#### Causal model

Now there is another measured variable `Z` which effects both `X` and `Y`. We might be only interested (for science or profit) in the causal relationship between `X` and `Y`, or the "direct effect" between `Z` and `Y`, or the "total effect" between `Z` and `Y`. We'll consider each of these below.

```{r}
n <- 500
Z <- rnorm(n)
X <- 5 -2 * Z + rnorm(n)
Y_CEF <- function(X, Z) 4 + 2 * Z - 3 * X
Y <- Y_CEF(X, Z) + rnorm(n)
```

Z: Cold weather (e.g., temperature)
X: Outdoor physical activity (e.g., frequency of outdoor exercise)
Y: Health service utilization (e.g., visits to hospitals)

Intuitively, Z negatively affects X and positively affects Y (in winter people are more likely to get flu). And X negatively affects Y.

```{r}
lm(Y ~ Z) |> broom::tidy() |> knitr::kable()
```

```{r}
lm(Y ~ X) |> broom::tidy() |> knitr::kable()
```

```{r}
lm(Y ~ X + Z) |> broom::tidy() |> knitr::kable()
```
#### Changing `X` only

```{r}
# Original data, copied for re-running
Z <- rnorm(n)
X <- 5 -2 * Z + rnorm(n)
Y <- Y_CEF(X, Z) + rnorm(n)
# Counterfactuals
Z_new <- Z
X_new <- X + 1
Y_new <- Y_CEF(X_new, Z_new) + rnorm(n)
```

#### Measuring the causal effect
The ATE is -3, because the coefficient for X is -3, and the intervention is 1.
```{r}
mean(Y_new - Y)
```

CATE at different values of `X`

```{r message=FALSE}
# Conditional average treatment effect (CATE)
qplot(X, Y_new - Y, geom=c("smooth", "point")) +
  geom_hline(yintercept = mean(Y_new - Y))
```

CATE at different values of `Z`

```{r message=FALSE}
# Conditional average treatment effect (CATE)
qplot(Z, Y_new - Y, geom=c("smooth", "point")) +
  geom_hline(yintercept = mean(Y_new - Y))
```

#### Changing `Z` but not `X` (direct effect)

Note that we are making an exception to our usual rule about propagating changes in the model. In this case we hold `X` fixed because we are interested in understanding only the "direct effect" of `Z` on `Y`, and not any change that would normally also go through `X`.

```{r}
# Original data, copied for re-running
Z <- rnorm(n)
X <- 5 -2 * Z + rnorm(n)
Y <- Y_CEF(X, Z) + rnorm(n)
# Counterfactuals
Z_new <- Z + 1
X_new <- X
Y_new <- Y_CEF(X_new, Z_new) + rnorm(n)
```
In this example we cut the link from Z to X, which usually violates the assumptions in causal inference.
#### Measuring the causal effect
```{r}
mean(Y_new - Y)
```
ATE is around 2, why?

CATE at different values of `X`

```{r message=FALSE}
# Conditional average treatment effect (CATE)
qplot(X, Y_new - Y, geom=c("smooth", "point")) +
  geom_hline(yintercept = mean(Y_new - Y))
```

CATE at different values of `Z`

```{r message=FALSE}
# Conditional average treatment effect (CATE)
qplot(Z, Y_new - Y, geom=c("smooth", "point")) +
  geom_hline(yintercept = mean(Y_new - Y))
```

#### Changing `Z` (total effect)


```{r}
# Original data, copied for re-running
Z <- rnorm(n)
X <- 5 -2 * Z + rnorm(n)
Y <- Y_CEF(X, Z) + rnorm(n)
# Counterfactuals
Z_new <- Z + 1
X_new <- 5 -2 * Z_new + rnorm(n)
Y_new <- Y_CEF(X_new, Z_new) + rnorm(n)
```

#### Measuring the causal effect

```{r}
mean(Y_new - Y)
```
ATE is around 8, why?


CATE at different values of `X`
All these CATE plots are nearly symmetric, why?
```{r}
qplot(X, Y_new - Y, geom=c("smooth", "point")) +
  geom_hline(yintercept = mean(Y_new - Y))
```
```{r}
qplot(Z, Y_new - Y, geom=c("smooth", "point")) +
  geom_hline(yintercept = mean(Y_new - Y))
```
Because in the whole domain, X and Z follow the same structural function Y, which means the treatment varies significantly across different subgroups of the population (defined by covariates X or Z). When subgroup varies significantly, CATE will be asymmetric, like what we did in the first example.

In above scenarios, we actually mimick RCTs: we know the true underlying function, then we add the intervention, followed by modelling the treatment. But what if RCTs are not available? Say we are given observations X and Y, and we dont know the true structural function, how can we infer that the real relationship between X and Y (which is Y=-3*X)? We need to infer the counterfactual relationship between X and Y given observations disrupted by unmeasured confounders. This is what causal inference scientists are researching and focusing. 

There are several methods, here I give an example of using a popular instrumental variable method 2SLS(2 stage least square) against the OLS(ordinary least square, usually linear regression).
```{r}
# Set seed for reproducibility
#install.packages("AER")
#library(AER)
set.seed(123)

n <- 500 

U <- rnorm(n, mean = 0, sd = 1)  # Unmeasured confounder

Z <- rnorm(n, mean = 0, sd = 1)  # Instrumental variable Z

X <- 0.8 * Z + 0.5 * U + rnorm(n, sd = 0.5)  # X depends on Z and U

Y <- -3 * X + 3 * U + rnorm(n, sd = 0.5)  # True causal effect of X on Y is -3, but U confounds the relationship

data <- data.frame(X, Y, Z)

ols_model <- lm(Y ~ X, data = data)
ols_predictions <- predict(ols_model)

iv_model <- ivreg(Y ~ X | Z, data = data)  
iv_predictions <- predict(iv_model)

Y_true <- -3 * X  # True relationship between X and Y without the confounder

plot_data <- data.frame(X = data$X, Y = data$Y, 
                        OLS_Pred = ols_predictions, 
                        IV_Pred = iv_predictions, 
                        True_Y = Y_true)

ggplot(plot_data, aes(x = X)) +
  geom_point(aes(y = Y), alpha = 0.6, color = 'black') + 
  geom_line(aes(y = True_Y, color = "True Structural Function", linetype = "True Structural Function"), size = 1) + 
  geom_line(aes(y = OLS_Pred, color = "OLS Prediction", linetype = "OLS Prediction"), size = 1) + 
  geom_line(aes(y = IV_Pred, color = "2SLS Prediction", linetype = "2SLS Prediction"), size = 1) +  
  labs(title = "True Function, OLS, and 2SLS Predictions",
       x = "X", y = "Y") +
  scale_color_manual(name = "Model", values = c("True Structural Function" = "green", 
                                                "OLS Prediction" = "red", 
                                                "2SLS Prediction" = "blue")) +
  scale_linetype_manual(name = "Model", values = c("True Structural Function" = "solid", 
                                                   "OLS Prediction" = "solid", 
                                                   "2SLS Prediction" = "solid")) +
  theme_minimal() +
  theme(legend.position = "top") +
  ggtitle("Comparison of True Structural Function, OLS, and 2SLS Predictions") +
  theme(plot.title = element_text(hjust = 0.5))

```
In above example we can see that with the help of instruments Z and the 2SLS algorithm, we successfully recover the causal relationship between X and Y, but if we run ordinary linear regression, it fails to give accurate predictions.

In the above example, we desing a perfect instrument that relates X, and it does not effect Y neither confounders. However, the search for such an instrument is very challenging in practice. In economic studies, the effort for seeking a good instrument is usually much more important than designing models.

Finding instruments: Economists, epidemiologists
Designing algorithms: Statisticians, machine learning scientists

**Note**: In this example the generating code is copy/pasted in 3 places, so any change you make you need to make the same change in all 3 code chunks. You might want to copy/paste the entire section and only edit the new copy of it.

- Try changing the functions/distributions generating each variable (but without altering the structure of the underlying DAG).

- Try to understand when regression coefficients give accurate answers, which causal effects they may be measuring, and when they do not give accurate answers.
